### 2022-11-30 chapter1/chapter1.tex
Made quotation marks consistent with LaTeX style

However, connected components and triangle clustering are special cases of the more general \textit{k}-clique percolation algorithm where \textit{k} equals two and three, respectively. The clique percolation algorithm detects clusters by first identifying cliques, sets of nodes which are fully connected, of a specified size \textit{k} in the graph (Fig.~\ref{fig:percolation}A). Clusters are then taken as the connected components of an overlap graph where an edge exists between two cliques if they share \textit{k}-1 nodes in common. An intuitive way to visualize this algorithm is by “rolling” a clique of some size \textit{k} over the graph (Fig.~\ref{fig:percolation}B). More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified. The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique. A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster. The strength of this algorithm is its ability to exclude sparsely connected nodes from clusters with an easily tunable parameter \textit{k}. Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships (Fig.~\ref{fig:percolation}C). We set \textit{k} to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).

@@

\subsection*{Alignment refinement}
Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long “poorly supported” segments which have no homology to most or any other sequences in the alignment. Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes “over-aligned” by forcing them into alignment where chance sequence similarities occur. Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions. More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment.

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, with higher values yielding a stronger correction. While values above 0.8 completely eliminate over-alignment and successfully align highly conserved regions, the alignment as a whole is severely degraded, as even homologous sequences with a small amount of divergence are separated by gaps. Thus, in our pipeline orthologous groups were aligned in two stages. In the first, the sequences were aligned with a strong correction of 0.7. Highly conserved regions were identified, which divided the alignment into a complementary set of diverged regions. The sequences in each of these regions were then extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4. The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment. By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}).

@@

Since the segments in these regions are likely the result of incorrect or incomplete annotations rather than meaningful biological variation, maintaining them in the alignments would propagate spurious homologies to subsequent analyses. This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or “trimming” columns below some threshold number of gaps or sequence identity~\cite{Castresana2000, CapellaGutierrez2009}. This approach, however, is inadequate if the taxonomic sampling is dense, as a single indel event along a lineage containing many species can increase the number of gaps above the threshold. Moreover, as this method does not incorporate any spatial information, it can rapidly alternate between trimming and preserving columns. Thus, it can severely disrupt any analyses which are sensitive to the spatial organization of an alignment.

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states”~\cite{Felsenstein1996}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, diverged regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively. To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left). The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by ``noise'' from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right). The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocess roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns. Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.

@@

Another challenge is the annotation of multiple isoforms for a single gene. Though prior pipelines have generally selected the longest isoform as the representative before conducting the orthology searches, if the sequences do not share a common intron-exon structure this approach can introduce artifacts or other issues during alignment. Instead, as protein sequences are increasingly derived from or linked to genomic sequences, we sought to incorporate the full annotations into the orthology inference pipeline. This, however, created two additional complications. First, a single gene could have several associated orthologous groups if its isoforms belonged to different clusters. Second, a single orthologous group could have several isoforms of a single gene if its isoforms were clustered together. In both cases, the presence of multiple sequences for a single gene creates ambiguities over which is the “primary” isoform. Since the first occurs at the level of orthologous groups, the orthologous groups were first grouped by the similarity of their parent genes using a graph-based strategy. Afterwards, a single representative is easily chosen as the group with the largest number of distinct species, though other criteria are possible. Since the second occurs within an orthologous group, the sequences were first aligned, and a representative for each gene was chosen as the sequence which was most concordant with this initial alignment.

@@

\subsection*{Merging of HSPs into hits}
HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs. In the first stage, proceeding from highest to lowest bit score, HSPs were marked as “disjoint” if they do not overlap with any other HSP previously marked as disjoint. Although this greedy strategy does not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs. In the second stage, all disjoint HSPs were marked as “compatible,” and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either. The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs. The best hits were then filtered by overlap and reciprocity criteria. The overlap criterion was applied first and requires that 50\% of residues in the query are aligned in compatible HSPs. This excludes false positives from conserved domains embedded in larger non-homologous proteins by ensuring the hits span a sufficient fraction of the query and target sequences. The reciprocity criterion requires each query-target pair has a corresponding hit where the roles are reversed, which ensures there is no ambiguity in which target is the best match for the query.

@@

The posterior distributions over states were computed for both the insertion and missing data phylo-HMMs. To trim the alignments, the posterior distributions from the insertion phylo-HMM were converted to state assignments with the following method. First the probabilities for states 2 and 3 were combined. Since this probability can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly. Instead, a high cutoff of 0.9 was used to define a “seed” region. The boundaries of the seed were then expanded both inwards and outwards to define intervals from which to select boundaries. The outward expansion halted when the probability or its derivative was below 0.05 and 0.001, respectively. The inward expansion halted when the derivative was below 0.02. The left and right boundaries were chosen as the columns in each expansion with the maximum product between the derivative and the number of sequences which changed to or from a gap between columns. By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries. In contrast, the posterior probabilities from the missing data phylo-HMM changed sharply, so positions were labeled as the ``missing data'' state if the posterior probability exceeded 0.95. These assignments can be used to filter segments or entire sequences from downstream analyses and are available in the supplementary data.

@@

To fit trees using the GTR model of nucleotide substitution, the protein alignments were converted to nucleotide alignments using the corresponding coding sequences in the genome annotations. Some protein sequences are “low quality,” meaning their coding sequences contain frameshifts, premature stop codons, or other errors even though they are strong hits to known protein-coding genes. The NCBI annotation pipeline corrects some of these defects in the protein sequences, which can complicate a simple “reverse translation” of the alignment. After rejecting alignments where the expected translation from a coding sequence differed from its corresponding protein sequence, 3,425 alignments remained. Consensus tree were derived from meta-alignments sampled from these alignments using the approach described previously except the GTR model was used in place of the LG model.

>>>>

However, connected components and triangle clustering are special cases of the more general \textit{k}-clique percolation algorithm where \textit{k} equals two and three, respectively. The clique percolation algorithm detects clusters by first identifying cliques, sets of nodes which are fully connected, of a specified size \textit{k} in the graph (Fig.~\ref{fig:percolation}A). Clusters are then taken as the connected components of an overlap graph where an edge exists between two cliques if they share \textit{k}-1 nodes in common. An intuitive way to visualize this algorithm is by ``rolling'' a clique of some size \textit{k} over the graph (Fig.~\ref{fig:percolation}B). More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified. The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique. A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster. The strength of this algorithm is its ability to exclude sparsely connected nodes from clusters with an easily tunable parameter \textit{k}. Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships (Fig.~\ref{fig:percolation}C). We set \textit{k} to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).

@@

\subsection*{Alignment refinement}
Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long ``poorly supported'' segments which have no homology to most or any other sequences in the alignment. Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes ``over-aligned'' by forcing them into alignment where chance sequence similarities occur. Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions. More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment.

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, with higher values yielding a stronger correction. While values above 0.8 completely eliminate over-alignment and successfully align highly conserved regions, the alignment as a whole is severely degraded, as even homologous sequences with a small amount of divergence are separated by gaps. Thus, in our pipeline orthologous groups were aligned in two stages. In the first, the sequences were aligned with a strong correction of 0.7. Highly conserved regions were identified, which divided the alignment into a complementary set of diverged regions. The sequences in each of these regions were then extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4. The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment. By defining highly conserved ``anchor'' regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}).

@@

Since the segments in these regions are likely the result of incorrect or incomplete annotations rather than meaningful biological variation, maintaining them in the alignments would propagate spurious homologies to subsequent analyses. This is a common issue in alignments generated by automated pipelines, so downstream analyses often focus on the strongly supported regions by removing or ``trimming'' columns below some threshold number of gaps or sequence identity~\cite{Castresana2000, CapellaGutierrez2009}. This approach, however, is inadequate if the taxonomic sampling is dense, as a single indel event along a lineage containing many species can increase the number of gaps above the threshold. Moreover, as this method does not incorporate any spatial information, it can rapidly alternate between trimming and preserving columns. Thus, it can severely disrupt any analyses which are sensitive to the spatial organization of an alignment.

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several ``hidden states''~\cite{Felsenstein1996}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, diverged regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively. To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left). The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by ``noise'' from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right). The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocess roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns. Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.

@@

Another challenge is the annotation of multiple isoforms for a single gene. Though prior pipelines have generally selected the longest isoform as the representative before conducting the orthology searches, if the sequences do not share a common intron-exon structure this approach can introduce artifacts or other issues during alignment. Instead, as protein sequences are increasingly derived from or linked to genomic sequences, we sought to incorporate the full annotations into the orthology inference pipeline. This, however, created two additional complications. First, a single gene could have several associated orthologous groups if its isoforms belonged to different clusters. Second, a single orthologous group could have several isoforms of a single gene if its isoforms were clustered together. In both cases, the presence of multiple sequences for a single gene creates ambiguities over which is the ``primary'' isoform. Since the first occurs at the level of orthologous groups, the orthologous groups were first grouped by the similarity of their parent genes using a graph-based strategy. Afterwards, a single representative is easily chosen as the group with the largest number of distinct species, though other criteria are possible. Since the second occurs within an orthologous group, the sequences were first aligned, and a representative for each gene was chosen as the sequence which was most concordant with this initial alignment.

@@

\subsection*{Merging of HSPs into hits}
HSPs were merged in two stages where the first combined non-overlapping HSPs, and the second combined the remaining HSPs. In the first stage, proceeding from highest to lowest bit score, HSPs were marked as ``disjoint'' if they do not overlap with any other HSP previously marked as disjoint. Although this greedy strategy does not necessarily yield the highest-scoring set of disjoint HSPs, it prioritized higher scoring HSPs. In the second stage, all disjoint HSPs were marked as ``compatible,'' and proceeding from highest to lowest bit score the remaining HSPs were marked as compatible if the overlap with any other compatible HSP was no more than 50\% of the length of either. The best hit for each query was chosen as the hit with the highest sum of bit scores from disjoint HSPs. The best hits were then filtered by overlap and reciprocity criteria. The overlap criterion was applied first and requires that 50\% of residues in the query are aligned in compatible HSPs. This excludes false positives from conserved domains embedded in larger non-homologous proteins by ensuring the hits span a sufficient fraction of the query and target sequences. The reciprocity criterion requires each query-target pair has a corresponding hit where the roles are reversed, which ensures there is no ambiguity in which target is the best match for the query.

@@

The posterior distributions over states were computed for both the insertion and missing data phylo-HMMs. To trim the alignments, the posterior distributions from the insertion phylo-HMM were converted to state assignments with the following method. First the probabilities for states 2 and 3 were combined. Since this probability can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly. Instead, a high cutoff of 0.9 was used to define a ``seed'' region. The boundaries of the seed were then expanded both inwards and outwards to define intervals from which to select boundaries. The outward expansion halted when the probability or its derivative was below 0.05 and 0.001, respectively. The inward expansion halted when the derivative was below 0.02. The left and right boundaries were chosen as the columns in each expansion with the maximum product between the derivative and the number of sequences which changed to or from a gap between columns. By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries. In contrast, the posterior probabilities from the missing data phylo-HMM changed sharply, so positions were labeled as the ``missing data'' state if the posterior probability exceeded 0.95. These assignments can be used to filter segments or entire sequences from downstream analyses and are available in the supplementary data.

@@

To fit trees using the GTR model of nucleotide substitution, the protein alignments were converted to nucleotide alignments using the corresponding coding sequences in the genome annotations. Some protein sequences are ``low quality,'' meaning their coding sequences contain frameshifts, premature stop codons, or other errors even though they are strong hits to known protein-coding genes. The NCBI annotation pipeline corrects some of these defects in the protein sequences, which can complicate a simple ``reverse translation'' of the alignment. After rejecting alignments where the expected translation from a coding sequence differed from its corresponding protein sequence, 3,425 alignments remained. Consensus tree were derived from meta-alignments sampled from these alignments using the approach described previously except the GTR model was used in place of the LG model.
